---
title: "Assignment 4"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2024-11-08"
---



```{r}
library(seqinr)
library(R.utils)
```


```{r, q1}
file_path <- file_path <- "C:\\Users\\moore\\OneDrive\\Desktop\\Fall 2024\\Programming for science informatics\\chr1_GL383518v1_alt.fa.gz"

gunzip(file_path, remove = FALSE)

decompressed_file_path <- gsub(".gz$", "", file_path)
```

```{r, task 1}
fasta_data <- read.fasta(decompressed_file_path)
# Access the 10th letter of the first sequence
tenth_letter <- fasta_data[[1]][10]
print(tenth_letter)
# Access the 758th letter of the first sequence
letter_758 <- fasta_data[[1]][758]
print(letter_758)
```

```{r, task 2}
sequence <- fasta_data[[1]]
reverse_complement <- rev(comp(sequence))

# Print the 79th letter of the reverse complement
print(paste("79th letter:", reverse_complement[79]))

# Print the 500th through the 800th letters of the reverse complement
subsequence_500_800 <- reverse_complement[500:800]
print("500th to 800th letters:")
print(subsequence_500_800)
```
```{r, task 3}
# Define kilobase size
kilobase_size <- 1000
sequence_length <- length(sequence)

# Initialize list to store counts per kilobase
nucleotide_counts_by_kb <- list()

# Loop through the sequence in 1000-base chunks
for (i in seq(1, sequence_length, by = kilobase_size)) {
  # Define the current kilobase range
  end_pos <- min(i + kilobase_size - 1, sequence_length)
  kilobase_chunk <- sequence[i:end_pos]
  
  # Count nucleotides in this chunk
  nucleotide_counts <- table(kilobase_chunk)
  
  # Ensure counts for each nucleotide, even if zero
  full_counts <- c(A = 0, T = 0, C = 0, G = 0)
  full_counts[names(nucleotide_counts)] <- nucleotide_counts
  
  # Store counts in list with kilobase number
  kb_number <- ceiling(i / kilobase_size)
  nucleotide_counts_by_kb[[paste0("Kilobase_", kb_number)]] <- full_counts
}

# Display the results
print(nucleotide_counts_by_kb)
```

```{r, task 4 part a}
# Convert the first kilobase counts to a data frame
first_kilobase_counts <- as.data.frame(t(nucleotide_counts_by_kb[[1]]))
colnames(first_kilobase_counts) <- c("A", "C", "G", "T")

# Display the data frame for the first 1000 base pairs
print(first_kilobase_counts)
```
```{r, task 4 part b}
# Combine all kilobase counts into a single data frame
all_kilobase_counts <- do.call(rbind, lapply(nucleotide_counts_by_kb, function(x) as.data.frame(t(x))))
colnames(all_kilobase_counts) <- c("A", "C", "G", "T")

# Display the first few rows of the data frame
print(head(all_kilobase_counts))
```
```{r, task 4 part c}
# This was done in 4b to ensure each row in the data frame corresponds to a kilobase
print(all_kilobase_counts)
```
```{r, task 4 part d}
# Calculate the sum for each kilobase row
all_kilobase_counts$Sum <- rowSums(all_kilobase_counts)

# Display the data frame with the sum column
print(head(all_kilobase_counts))
```

1. The expected sum for each list is 1000.
```{r}
expected_sum <- 1000
# Identify rows where the sum does not match the expected 1000 bases
anomalies <- all_kilobase_counts$Sum != expected_sum

# Display rows with discrepancies, if any
if (any(anomalies)) {
  print("Rows with sums not equal to the expected value of 1000:")
  print(all_kilobase_counts[anomalies, ])
} else {
  print("All rows have the expected sum of 1000.")
}

```
2. Yes the sum of 439 in Kilobase_183 confirms that the last part of the sequence is shorter than 1000 base pairs. 

3. Explanation for discrepancies: The primary reason for any discrepancy in the sum is likely due to the last kilobase in the sequence, which may not contain a full 1000 bases if the total sequence length is not a perfect multiple of 1000. This would result in a sum less than 1000 (439) for the last kilobase.





